Working with Daily Climate Model Output Data in R and the \pkg{futureheatwaves} Package

### 2017-05-23 paper_drafts/RJournalRevisions2/response_to_reviewers.txt

Thank you very much for the opportunity to revise and resubmit this manuscript. The
reviewers made some excellent suggestions, and we have modified the manuscript to
respond to these. In cases where suggested changes are beyond the scope of the described
package, we have added text to the manuscript to explain how such proposed approaches could
contribute as R programmers continue to evolve software solutions to working with climate model
output in R, and we have tried to make some connections to how R-based approaches to working with
large data continue to evolve in another topical areas with large, array-based data (flow
cytometry). In addition, we have revised and updated the package that is described by this
manuscript, both in response to reviewer suggestions and also to ensure functionality with recent
updates to `dplyr` and other key imported packages that have undergone major updates since the
original submission.

Comments from Reviewer #1

Comment: The authors answered quite satisfactory to the questions and also greatly
improved the manuscript and the package. I recommend that the article
should be published "as is" after solving a minor issue of the package: the
function 'map_grid_leaflet' does not work correctly because it requires a
local copy of the leaflet library.  Is it possible to put the http link of
leaflet instead of a local one so that function works for any connected
machine?

Response: The current version of the package imports the latest version of
`leaflet` from CRAN (version 1.1.0) and works with this version of leaflet,
rather than a local copy of the leaflet library. The function will now
work for any connected machine.

... https://github.com/rstudio/leaflet/issues/192#issuecomment-286883019
Also, "a network proxy which would block https requests" can cause problems?
"Do you have a network proxy or VPN active on your machine or your network?"
RStudio advice: https://support.rstudio.com/hc/en-us/articles/200488488-Configuring-R-to-Use-an-HTTP-or-HTTPS-Proxy

Comments from Reviewer #2

Comment: Climate science is fundamental to informing key societal decisions in the 21st
century. The Climate Intercomparison Project (currently in Phase 5) is a valuable
suite of Early system model outputs that continues to produce novel science. As
such, R packages to ingest and analyze these outputs could be quite valuable to the
community, particularly to field scientists and experimentalists who are not familiar
with the more idiosyncratic pipelines traditionally used by Earth system modelers. That
being said, this package does not address the most pressing barriers to data access and
interpretation, and I'm unclear that this manuscript describes a package that is
sufficiently novel to warrent publication.

Response: We appreciate the reviewer's point that the described package does not
resolve all the challenges of working with climate model output data in R.
However, we think that the package described offers an important step forward in making the
ensemble techniques that have become popular in climate science research more accessible to
researchers who are conducting climate impact studies but are not themselves climate scientists.
We have added a "Motivation" subsection to the manuscript section describing the `futureheatwaves`
package to clarify this point:

Added text: "We created the `futureheatwaves` package to aid in identifying, characterizing, and
exploring multi-day extreme events in daily climate model output data. While most of the discrete
tasks involved in identifying and characterizing multi-day extreme events are fairly
straightforward, the full process can be code-intensive, especially for multi-city studies and
studies that test sensitivity to how an event is defined or that incorporate different scenarios of
adaptation in the case of events defined using a threshold relative to community climate. Our aim
in developing this package was therefore to make the full process of identifying and characterizing
these extreme events much more convenient and so facilitate the use of multi-model, multi-ensemble
member analyses in climate impact studies conducted by non-climate scientists."

Further, we have also added a section to the end of the paper ("Future directions for working
with climate model output in R") that describes how future software development in R could
tackle the challenges provided by climate data, to help put the package described in the current
manuscript in context within the evolving use of R for research involving climate model output
data.

Comment: This article correctly identifies memory use and data format as a key issue for the
usability of CMIP5 data, neither of which are addressed by this package. The user is required
to use `ncdf4` to preprocess the CMIP5 outputs and script their own sub-setting routines. This
sidesteps both the memory issue and data formatting. The comments on calendars being non-standard
in the CMIP5 models is, again, useful and accurately identified but not addressed systematically
in the package and left as an issue for the user to solve, albeit, with guidance. If parts of this
process were hard coded (for example, Model A has a 365-day calendar starting from 1800 and model
B has a leap-year calendar starting from 1900) this could be very useful to the community.

Response: We appreciate the reviewer's point that having information at hand on which calendar
a climate models uses might be useful to those working with climate model output in R. Such a
table, however, would be fairly complex if it included both the calendar and the reference date
from which days are counted for all CMIP5 data. This is because, for some
of the CMIP5 climate models, different calendars are used for different
experiments, and difference reference dates are also used for different experiments within climate
models.

To provide some guidance to R users, however, we have added a link to the manuscript for a table
provided in the Frequently Asked Questions section of the Earth System CoG website. This table
gives the calendar (although not all starting reference dates) for each of the CMIP5 models,
including notes of exceptions for certain experiments within some models. Here is the sentence
we added to the manuscript with this link:

Added text: "While most CMIP5 climate models use the same calendar for all experiments, a few do not; a full table of the calendars used for each climate model and experiment, pulled from netCDF metadata, is available at https://www.earthsystemcog.org/projects/cog/faq_data."

In terms of the challenges presented by the large file sizes of climate
model output, we do recommend some pre-processing of data files to subset files down
to the geographic area and time range required for a study (e.g., North America for a national
study of the United States) which would help reduce data size prior to processing climate model
output through the described package. As the reviewer notes, we offer suggestions and guidance for
conducting this pre-processing in R, both through text in the manuscript. We offer additional
guidance through a vignette provided with the package described in the manuscript
(https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html).
Even after pre-processing the files prior to using the `futureheatwaves` package, many research
applications will still involve working with very large files. For example, a national study of the
United States may still use files with daily temperatures for many decades at a large number of
grid points, and an analysis that uses multiple climate models, some with multiple ensemble
members, would still need to process a number of very large files. The `futureheatwaves` package
addresses these large file sizes by inputting and processing data from each ensemble member of each
climate model separately, avoiding the memory problems created by trying to load all data from all
ensemble members into memory at once and processing the data using `apply`, `map`, `plyr`, or
similar tools applied across all data at once.

We acknowledge, however, that while this approach is successful in terms of allowing the
identification and characterization of sustained extreme events in R, future software development
could continue refining and optimizing the handling of climate model data in R. We have added a
section to the end of the manuscript on future directions for climate model output data in R and
have included comments on how such future work might be able to successfully draw on CDF-backed
approaches recently created for handling flow cytometry data in R or on "data vault"-style
approaches geared to improve ...

[Added text]

Comment: Some acknowledgement of NCL or CDO for merging, sub-setting, averaging, and otherwise
manipulating netCDF files is warranted in this manuscript. These processes are MUCH faster in
these specialty languages and then done via R and not to mention this when dealing with CMIP5
data is remiss. One idea, although this completely changes the focus of the package, would be
to create a wrapper for these speciality languages in R to leverage that speed.

Response: Thank you for this suggestion. In the revised version of the manuscript, we have
added references to both software tools, as well as text describing the availability of
NCL and CDO for initial processing of climate model output files, including both the computation speed advantages and the disadvantage of needing to learn the syntax of a new language:

"CMIP5 data are available in the netCDF file format. Free specialty software exists to work with
climate model output files in this format, including a collection of command line tools developed
by the Max Planck Institute called Climate Model Operators (CDO) (Schulzweida, 2017) and an
interpreted language developed by the National Center for Atmospheric Research (NCAR) called the
NCAR Command Language (NCL) (UCAR/NCAR Computational and Information SystemsLaboratory, 2017).
Although such software can be used to quickly process netCDF climate model output files, they
require learning a new language syntax, and so for R users may not be worth the computational speed
gain compared to alternative solutions that can be scripted in the R language."

Further, we have added a section to the end of the paper, in response to this and other comments
from this reviewer, called "Future directions for working with climate model output in R". In this
section, we have added some text outlining the reviewer's idea that wrappers for these two
speciality languages could be an interesting next step to pursue as R software for working with
climate model output data evolves, including references to existing libraries with wrappers in
other languages (Python and Ruby):

Added text: "One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available
through the Climate Data Operators (CDO) software (Schulzweida, 2017). Libraries already exist for
Python and Ruby that allow the functionality of CDO tools to be used within these scripting
languages (available from the Max Planck Institute at
https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D). While one R package (`ncdf4.helpers`;
Bronaugh, 2014) already provides R wrappers for a few CDO operators, such functionality could be
extended through future R software to capture more of the full functionality of the CDO toolkit."

Comment: The analysis that this package does is fairly straightforward using `plyr` or other
common R packages / functions. The main function described in this article is the identifying
and summing ranges of events. If a user made it far enough to do the pre-processing of the CMIP5
output data I doubt this would be a computational barrier to further analysis. If authors could
show notable speed or memory requirement improvement then this could be interesting.

Response:

Comment: Readability of the article is excellent.

Response: Thank you. We have tried to maintain the same level of readability through our revisions
in response to reviewers' comments.

### 2017-05-23 paper_drafts/RJournalRevisions2/anderson-eason-barnes.Rmd
Work on revisions for R Journal paper

CMIP5 data are available as netCDF files, for which base R import functions do not exist. There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \CRANpkg{ncdf4} \citep{CRANncdf4} and \pkg{CRANRNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.

@@

Two R packages provide help with non-Gregorian calendars: \CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers} \citep{ncdf4.helpers}. The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class. This class is defined in the \pkg{PCICt} package and provides date-like functionality for 360- and 365-day calendars \citep{PCICt}. However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name. The \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

@@

@@

We created the \pkg{futureheatwaves} package to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data. Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}). This package start from files rather than R objects to avoid loading data from all climate model ensembles at once; instead, the function loads, processes, and saves output for a single climate model ensemble member at a time. The extreme events are identified and characterized at one or more study locations (e.g., cities), which the user specifies in an input file. The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets").

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
\end{center}
\caption{Snapshot of an interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}. The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used. The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example. From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
\label{fig:gridmap}
\end{figure}

@@

>>>>

CMIP5 data are available in the netCDF file format. Free specialty software exists to work with climate model output files in this format, including a collection of command line tools developed by the Max Planck Institute called Climate Model Operators (CDO) \citep{schulzweida2006cdo} and an interpreted language developed by the National Center for Atmospheric Research (NCAR) called the NCAR Command Language (NCL) \citep{ncl}. Although such software can be used to quickly process netCDF climate model output files, they require learning a new language syntax, and so for R users may not be worth the computational speed gain compared to alternative solutions that can be scripted in the R language. While base R import functions do not exist for netCDF files, there are a few R package extensions that allow R users to work with the netCDF file format used for CMIP5 files directly from R. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \CRANpkg{ncdf4} \citep{CRANncdf4} and \pkg{CRANRNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.

@@

Two R packages provide help with non-Gregorian calendars: \CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers} \citep{ncdf4.helpers}. The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class. This class is defined in the \pkg{PCICt} package and provides date-like functionality for 360- and 365-day calendars \citep{PCICt}. However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be successfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R. While most CMIP5 climate models use the same calendar for all experiments, a few do not; a full table of the calendars used for each climate model and experiment, pulled from netCDF metadata, is available at \url{https://www.earthsystemcog.org/projects/cog/faq_data}.

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the file names for all files in a directory to extract information within the file name. For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name. The \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

@@

## Motivation

We created the \pkg{futureheatwaves} package to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data. While most of the discrete tasks involved in identifying and characterizing multi-day extreme events are fairly straightforward, the full process can be code-intensive, especially for multi-city studies and studies that test sensitivity to how an event is defined or that incorporate different scenarios of adaptation in the case of events defined using a threshold relative to community climate. Our aim in developing this package was therefore to make the full process of identifying and characterizing these extreme events much more convenient and so facilitate the use of multi-model, multi-ensemble member analyses in climate impact studies conducted by non-climate scientists.

@@

Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}). This package start from files rather than R objects to avoid loading data from all climate model ensembles at once; instead, the function loads, processes, and saves output for a single climate model ensemble member at a time. The extreme events are identified and characterized at one or more study locations (e.g., cities), which the user specifies in an input file. The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets").

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
\end{center}
\caption{Snapshot of an interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}. The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used. The interactive maps include pop-ups with city identifiers; one is shown open in this snapshot as an example. From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
\label{fig:gridmap}
\end{figure}

@@

# Future directions for working with climate model output in R

Research that assesses the potential impacts of climate change is critical in informing current policy choices, and R is an important tool for many researchers performing such assessments. While the `futureheatwaves` package described here takes steps to facilitate the assessment of impacts related to sustained, multi-day events, a number of challenges remain in working with climate model output data in R, and future R software development offers the potential to further address the challenges of working with this data.

One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available through the Climate Data Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already exist for Python and Ruby that allow the functionality of CDO tools to be used within these scripting languages (available from the Max Planck Institute at \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.

Another important path for development could be through approaches that allow researchers to take advantage of the statistical tools offered by R while maintaining large climate model output files in a netCDF format. For example, Goncalves and coauthors recently described a ``round table" approach of connecting as-needed data access from netCDF climate data files through to functionality available in R and CDO through the intermediary of a MonetDB database system \citep{goncalves2015round}. In other topical areas, R programmers are also improving the efficiency of working with data in large netCDF files through approaches that avoid loading all data in-memory. For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{}. Such an approach could be promising for future R software development for working with large climate model files.

### 2017-05-23 paper_drafts/RJournalRevisions2/anderson-eason-barnes.tex
Work on revisions for R Journal paper

Two R packages provide help with non-Gregorian calendars:
\CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers}
\citep{ncdf4.helpers}. The \code{nc.get.time.series} function in
\pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in
the CMIP5 netCDF file's attributes to convert the ``time'' variable in
the file to an object of the \code{PCICt} class. This class is defined
in the \pkg{PCICt} package and provides date-like functionality for 360-
and 365-day calendars \citep{PCICt}. However, while these functions will
help with handling most CMIP5 files, the CMIP5 standards allows use of
other calendars which may not be succesfully handled by these functions,
so it is important to assess whether the time variable range in the
\code{PCICt} object correctly matches the expected date ranges for a
file when processing CMIP5 data in R.

@@

@@

We created the \pkg{futureheatwaves} package to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). This package start from files rather than
R objects to avoid loading data from all climate model ensembles at
once; instead, the function loads, processes, and saves output for a
single climate model ensemble member at a time. The extreme events are
identified and characterized at one or more study locations (e.g.,
cities), which the user specifies in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``).

@@

\begin{Schunk}
\begin{Sinput}
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

@@

>>>>

Two R packages provide help with non-Gregorian calendars:
\CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers}
\citep{ncdf4.helpers}. The \code{nc.get.time.series} function in
\pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in
the CMIP5 netCDF file's attributes to convert the ``time'' variable in
the file to an object of the \code{PCICt} class. This class is defined
in the \pkg{PCICt} package and provides date-like functionality for 360-
and 365-day calendars \citep{PCICt}. However, while these functions will
help with handling most CMIP5 files, the CMIP5 standards allows use of
other calendars which may not be successfully handled by these
functions, so it is important to assess whether the time variable range
in the \code{PCICt} object correctly matches the expected date ranges
for a file when processing CMIP5 data in R. While most CMIP5 climate
models use the same calendar for all experiments, a few do not; a full
table of the calendars used for each climate model and experiment,
pulled from netCDF metadata, is available at
\url{https://www.earthsystemcog.org/projects/cog/faq_data}.

@@

\subsection{Motivation}\label{motivation}

We created the \pkg{futureheatwaves} package to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. While most of the discrete tasks involved in
identifying and characterizing multi-day extreme events are fairly
straightforward, the full process can be code-intensive, especially for
multi-city studies and studies that test sensitivity to how an event is
defined or that incorporate different scenarios of adaptation in the
case of events defined using a threshold relative to community climate.
Our aim in developing this package was therefore to make the full
process of identifying and characterizing these extreme events much more
convenient and so facilitate the use of multi-model, multi-ensemble
member analyses in climate impact studies conducted by non-climate
scientists.

@@

Figure \ref{fig:overview} gives an overview of the two primary functions
of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set}
function processes a directory of climate projection files that are
stored locally on the user's computer (Figure \ref{fig:overview},
``Climate projections''), to generate a list of all extreme events in
each projection, as well as over a dozen characteristics of each
identified extreme event (Table \ref{tab:hwcharacteristics}). This
package start from files rather than R objects to avoid loading data
from all climate model ensembles at once; instead, the function loads,
processes, and saves output for a single climate model ensemble member
at a time. The extreme events are identified and characterized at one or
more study locations (e.g., cities), which the user specifies in an
input file. The extreme events identified for each ensemble member are
output as separate files in a directory specified by the user (Figure
\ref{fig:overview}, ``Extreme events datasets``).

@@

\begin{Schunk}
\begin{Sinput}
library(futureheatwaves)
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

@@

\section{Future directions for working with climate model output in
R}\label{future-directions-for-working-with-climate-model-output-in-r}

Research that assesses the potential impacts of climate change is
critical in informing current policy choices, and R is an important tool
for many researchers performing such assessments. While the
\texttt{futureheatwaves} package described here takes steps to
facilitate the assessment of impacts related to sustained, multi-day
events, a number of challenges remain in working with climate model
output data in R, and future R software development offers the potential
to further address the challenges of working with this data.

One important step in future development of R software to work with
climate model output could be the development of R wrappers for some of
the existing command line tools available through the Climate Data
Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already
exist for Python and Ruby that allow the functionality of CDO tools to
be used within these scripting languages (available from the Max Planck
Institute at
\url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R
package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already
provides R wrappers for a few CDO operators, such functionality could be
extended through future R software to capture more of the full
functionality of the CDO toolkit.

Another important path for development could be through approaches that
allow researchers to take advantage of the statistical tools offered by
R while maintaining large climate model output files in a netCDF format.
For example, Goncalves and coauthors recently described a ``round table"
approach of connecting as-needed data access from netCDF climate data
files through to functionality available in R and CDO through the
intermediary of a MonetDB database system \citep{goncalves2015round}.

In other topical areas, R programmers are also improving the efficiency
of working with data in large netCDF files through approaches that work
with the data on-disc {[}?{]}. For example, the recent Bioconductor
package \pkg{openCyto} {[}?{]} allows scientists to work with the large
files generated by flow cytometry that characterize cell populations in
a sample. Similar approaches have been taken in metabolomics (e.g.,
{[}packages{]}) and other 'omics fields. The developers of future R
packages to work with climate model output may find it useful to borrow
ideas from these fields to create solutions with improved memory
management {[}?{]} for large climate output files.
